Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2024-10-30 docs/paper.tex
Include regime difference test results table

difference being statistically significant for one commodity. The
equally weighted US portfolio had a mean return of 15.7\% over
\textbf{backwardation} against 16.2\% during phases of
\textbf{contango}, both statistically significant at the 10\% level. The
difference was considerably more pronounced for the equally weighted GB
metals portfolio with a mean return of 8\% during \textbf{backwardation}
rising to 35.6\% during \textbf{contango}.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities; this sweeping impact spurred the transformation of
commodity markets from a collection of idiosyncratic markets into a more
homogeneous collection akin to an asset class. In this context
interconnectedness emerges as a crucial aspect that warrants thorough
examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicates higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We further examine these correlations by year in
Table~\ref{tbl-correlations-years}. For the full sample of assets, the
average pairwise correlations remain in the 0.06--0.08 range from 1998
through 2003, before suddenly increasing to 0.12 from 2003 to 2004,
suggesting 2004 as a distinct break point corresponding to the onset of
financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

traded commodities considered in the study with equal weighting. This
portfolio enables a refined examination of cross-correlations across
sectors as the numerator of the beta of the regression of the returns
for an individual commodity against those for the market factor is
driven by covariances while the denominator is driven by the individual
commodity variance. Thus, an increase in a commodity's correlation with
other commodities should lead to an increase in its beta relative to the
market factor portfolio. We accordingly regress the weekly individual
commodity front futures returns against those for our market factor
portfolio and average the resulting betas across the entire set of
commodity assets considered as well as by country, sector and sub-sector
for each of the four periods of interest with the results shown in

@@

precious sub-sector in particular. The average \(R^{2}\) for the group
jumps to 55\% during financialisation from less than 12\% over the past
period while the short leg sees a far smaller increase, up to 8.9\% from
3.3\% over the past period. Interestingly, although the term structure
factor itself only sees a very modest increase, its short leg (low
roll-yield assets: Working's backwardation) shows a sizeable rise in
pricing performance, yet again driven the metals sector, with the
average \(R^{2}\) for the group's US traded assets jumping to 28.2\%
during financialisation from 6.2\% over the past period. The pattern of
results for both factors extends to the GB traded assets and suggest a
convergence of Keneysian and Working's term structure perspective over

@@

of financialisation on the metals complex and provide further evidence
that global metals co-movement could be detected by US based factor
models that combine the Keynesian and Working's term structure
paradigms. There seems thus to be information in the factor constructed
from hedging pressure based on the Keynesian paradigm relevant to the
term structure factor built on the Working paradigm and these findings
therefore appear to complement the results of \citet{kang_tale_2020} who
find that hedging pressure both conveys information about liquidity
provision in the short term (Working paradigm) and risk transfer in the
longer term (Keynesian paradigm).

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-combined}) with eighteen US commodities along with
all six GB traded metals showing lower mean returns. Returns for equally
weighted portfolios of both sets of assets fall down to 9.4\% and 5.3\%
from 15.9\% and 21.8\% respectively. Returns are mostly higher during
phases of \textbf{backwardation} relative to \textbf{contango} with
fifteen US assets and five of the GB traded metals showing higher mean
returns in \textbf{backwardation} and the pattern extending to equally
weighted portfolios of the two sets of assets. Volatility is moderately
higher for both sets with fifteen US commodities and three GB metals
showing higher figures and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks in the crisis period (Table~\ref{tbl-time-reg-US-CHP}) with both
now seemingly converging; the US cross-sectional average \(R^{2}\) at
the individual level bouncing back up (22.5\% vs 20.8\%) while that at
the aggregate level rises further (14.7\% vs 7.1\%).

The average pairwise correlation for the entire set of sample assets
increases further into the crisis to 0.28
(Table~\ref{tbl-correlations-periods}) indicating an intensification of
the higher interdependence within the full cross-section of commodities
prompted by financialisation. The result is consistent with the greater
interdependence between equity and commodity markets over this period
documented in \citet{silvennoinen_financialisation_2013}. The figure for
the US traded section of the metals sector only progresses modestly to
0.59 after peaking at 0.63 in 2006 (Table~\ref{tbl-correlations-years}).
In contrast, the GB traded assets see a sharp increase to over 0.65 in
the crisis starting in 2007 up to a peak of 0.69 in 2010, and so does
the agricultural sector with the sharpest increase in 2008 when the
average pairwise correlation for the sector jumps to 0.30. The energy
sector also shows a substantial progression starting in 2007 and peaking
at 0.63 in 2008 before levelling back down over the rest of the period.
The onset of the financial crisis seems therefore to have had an across
the board impact on the interdependence of commodities in contrast with
financialisation, much more closely related to the metals sector.

The entire cross-section sees an increase in average regression
coefficient relative to the market factor over the crisis period with
the strongest progression for the GB traded metals followed by the
agricultural sector (Table~\ref{tbl-regressions-index}). The pricing
performance of the factor dominates the US commodity complex over the
period (Table~\ref{tbl-regressions-factors}), showing a 50\% increase in
mean average \(R^{2}\) to over 31\%, while that for both the CHP and
open interest factor decreases and, although increasing, remains modest
for the term structure factors; the increase is strongest for the GB
traded metal assets with an average \(R^{2}\) of almost 34\% for the
period. Taken together, these results indicate the presence of a
systematic factor across the entire cross-section of the US commodities.
For both the CHP and term structure factors, both legs show an increase
in average \(R^{2}\) and while substantially diverging over the
preceding financialisation period the pricing performance for both legs
seems to converge back to a pattern closer to that observed
pre-financialisation.

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-combined}) with twenty-two US and four GB traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all GB metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and GB assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all GB metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and GB assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period with mean returns higher during phases of
\textbf{backwardation} relative to \textbf{contango}, more so than over
the crisis period with seventeen US commodities showing higher mean
returns over \textbf{backwardation} phases. The difference is
significant at the 5\% level for Lumber (XCME) and at the 10\% level for
Palladium (XNYM) while no difference is significant for the assets
showing higher figures over phases of \textbf{contango}. The pattern is
more pronounced for the GB traded subset with all six assets showing
higher mean returns in \textbf{backwardation} with the difference
significant at the 1\% level for Nickel-primary (XLME), 5\% level for
Aluminium (XLME) and 10\% level for Zinc (XLME). It also extends to
equally weighted portfolios of US and GB traded assets that both show
higher mean returns during \textbf{backwardation} with the difference
significant at the 5\% level for the GB portfolio and the figure close
to the 10\% level significance threshold for the portfolio of US assets.
Volatilities are overall higher in \textbf{contango} although less so
than over the previous period with seventeen US commodities showing
higher figures over \textbf{contango} against twenty-three in the crisis
period. The pattern extends to three of the six GB traded metals with
the difference significant at the 1\% level for Copper (XLME) and Tin
(XLME) and at the 5\% level for Zinc (XLME) while differences are not
significant for Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME)
that exhibit higher volatility over \textbf{backwardation} phases.
Equally weighted portfolios of US and GB traded assets on the other hand
both show higher volatility in \textbf{contango} with the difference
significant at the 1\% level for both. Over the financialisation period
in contrast volatilities were predominantly higher in
\textbf{backwardation} for US traded assets with fourteen commodities
showing higher volatility figures in these phases; the post-crisis
pattern of results for volatilities thus likewise shows signs of
reversion to pre-financialisation standards. After showing signs of
convergence in the crisis period the dynamics of the explanatory power
of CHP and \textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}),
seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a return to legacy commodity
specific dynamics. The recovery initiated in the crisis period for that
of CHP is confirmed with a stronger increase in the post-crisis period,
27.2\% vs.~22.5\%, strongest for the metals sector, 37.3\% vs.~23.7\%,
while the energy sector shows the opposite pattern: 11\% vs.~16\%. In
contrast, after increasing stronger in the crisis, that for \textbf{CHP}
falls overall in the post-crisis period, 7.5\% vs 14.7\%, with the
decrease strongest for the energy sector in a relative sense: 2.5\%
vs.~14.8\%.

The average pairwise correlations fall across the board in the
post-crisis period, below the levels observed during financialisation,
with the figure for the entire cross-section
(Table~\ref{tbl-correlations-periods}) dropping down to 0.12 from 0.28
over the crisis. This substantial decline, that does not appear to have
been widely noted, appears to begin in the latter part of the crisis
period, around 2012 (Table~\ref{tbl-correlations-years}), before
levelling down over the whole post-crisis period.

Similarly, the average market factor beta for the entire cross-section
drops down to 0.94 (Table~\ref{tbl-regressions-index}), below the level
observed during financialization. The decline seems driven by the metals
sector, particularly the GB traded segment, with the figures for the US
and GB traded assets at 0.91 and 0.72, down from 1.12 and 1.18,
respectively. Meanwhile, the pricing performance of the market factor
also retreats to below financialisation levels with the average
\(R^{2}\) for the entire cross-section dropping down to 14.7\%
(Table~\ref{tbl-regressions-factors}). All sectors seem affected with
agriculturals, energy and metals all seeing their average \(R^{2}\)
decline below the levels observed during the financialisation period
albeit remaining above 10\% for the GB traded metals against less than
2\% pre-financialisation. Similarly, the performance for both the CHP
and term structure factors falls back to levels below those attained
over financialisation while the gap between their long and short legs
remains narrower than observed during that period. These results, along
with those from the crisis period, suggest that the exogenous shock of
the financial crisis, combined with the subsequent accommodative
monetary policy, may have triggered a reversion toward
pre-financialization fundamentals in the commodity futures complex; an
issue that, to our knowledge, has not been widely discussed in the
existing literature.\\
The phases analysis further documents the emerging role/dominance of the
\textbf{contango} regime post-crisis with average correlations higher
during those phases and the pattern stronger for the metal assets both
US and GB traded. Similarly, for the latter the pricing performance of
the market factor is considerably better during phase of
\textbf{contango} in contrast with the previous periods that showed no
substantial difference between phase of \textbf{contango} and
\textbf{backwardation}. The figures for the US and GB traded segments of
the sector are 23\% (\textbf{contango}) vs 11.5\%
(\textbf{backwardation}) against 36.2\% vs 43\% during financialisation
and 14.4\% (\textbf{contango}) vs 6.6\% (\textbf{backwardation}) against
33.5\% vs 34.4\% during financialisation respectively. Taken together,
these results indicate some recent changes in the correlation structure
within the sector and along with the decline in interdependence in the
commodity sector at large documented above, this observation could form
a valuable basis for future research.\\
These findings providence additional evidence that the interdependence
effects of financialization seems to have been strongest for the entire
cross-section of the metals sector with a lesser impact for the other
sectors.

@@

>>>>

difference being statistically significant at the 5\% level for one
commodity (Nickel-primary (XLME)). The equally weighted US portfolio had
a mean return of 15.7\% over \textbf{backwardation} against 16.2\%
during phases of \textbf{contango}, both statistically significant at
the 10\% level. The difference was considerably more pronounced for the
equally weighted GB metals portfolio with a mean return of 8\% during
\textbf{backwardation} rising to 35.6\% during \textbf{contango}.

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more homogeneous
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average of weekly
returns pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector. Table~\ref{tbl-correlations-periods}
shows the results for the four periods of interest with the results
reported for the entire period as well as for periods of
\textbf{backwardation} and \textbf{contango}. For the entire sample the
average pairwise correlation jumps from 0.07 in the first period to 0.16
over financialisation. This finding indicates higher interdependence
within the full cross-section of commodities starting with the
financialisation period, consistent with the empirical findings in
\citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods which indicates
that increased commodity interdependence over financialisation was not a
phase driven phenomenon in the same way that individual commodity
returns are, as demonstrated in Table~\ref{tbl-stats-combined}.\\
We break down these average correlations by years in
Table~\ref{tbl-correlations-years}. For the full sample of assets, the
average pairwise correlations remain in the 0.06--0.08 range from 1998
through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This
suggests 2004 as a distinct break point for the onset of
financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

traded commodities with equal weighting. This portfolio enables a
refined examination of cross-correlations across sectors as the
numerator of the beta of the regression of the returns for an individual
commodity against those for the market factor is driven by covariances
while the denominator is driven by the individual commodity variance.
Thus, an increase in a commodity's correlation with other commodities
should lead to an increase in its beta relative to the market factor
portfolio. We accordingly regress the weekly individual commodity front
futures returns against those for our market factor portfolio and
average the resulting betas across the entire set of commodity assets
considered as well as by country, sector and sub-sector for each of the
four periods of interest with the results shown in

@@

precious sub-sector in particular. Interestingly, although the term
structure factor itself only sees a very modest increase, its short leg
(low roll-yield assets: Working's backwardation) shows a sizeable rise
in pricing performance, yet again driven the metals sector. The pattern
of results for both factors extends to the GB traded assets and suggest
a convergence of Keneysian and Working's term structure perspective over

@@

of financialisation on the metals complex suggest that there is
information in the factor constructed from hedging pressure based on the
Keynesian paradigm relevant to the term structure factor built on the
Working paradigm. These findings therefore appear to complement the
results of \citet{kang_tale_2020} who find that hedging pressure both
conveys information about liquidity provision in the short term (Working
paradigm) and risk transfer in the longer term (Keynesian paradigm).

@@

The crisis period (late 2008 to mid-2013) shows a contrasting pattern of
results. Returns fall sharply relative to the financialisation period
(Table~\ref{tbl-stats-combined}) with eighteen US commodities along with
all six GB traded metals showing lower mean returns. Returns for equally
weighted portfolios of both sets of assets fall down to 9.4\% and 5.3\%
from 15.9\% and 21.8\% respectively. Returns are mostly higher during
phases of \textbf{backwardation} relative to \textbf{contango} with
fifteen US assets and five of the GB traded metals showing higher mean
returns in \textbf{backwardation} and the pattern extending to equally
weighted portfolios of the two sets of assets. Volatility is moderately
higher for both sets with fifteen US commodities and three GB metals
showing higher figures and interestingly eminently higher over phases of
\textbf{contango}; out of the US complex only Sugar-\#11 (IFUS) and none
of the GB traded metals show higher volatility figures over
\textbf{backwardation} phases. The symmetry between the dynamics of CHP
and \textbf{CHP}'s explanatory power on individual commodity returns
breaks (Table~\ref{tbl-time-reg-US-CHP}) with both now seemingly
converging; the US cross-sectional average \(R^{2}\) at the individual
level bouncing back up (22.5\% vs 20.8\%) while that at the aggregate
level rises further (14.7\% vs 7.1\%).

The average pairwise correlation for the entire set of sample assets
increases further to 0.28 (Table~\ref{tbl-correlations-periods})
indicating an intensification of the higher interdependence within the
full cross-section of commodities prompted by financialisation. The
result is consistent with the greater interdependence between equity and
commodity markets over this period documented in
\citet{silvennoinen_financialisation_2013}. The figure for the US traded
section of the metals sector only progresses modestly to 0.59 after
peaking at 0.63 in 2006 (Table~\ref{tbl-correlations-years}). In
contrast, the GB traded assets see a sharp increase to over 0.65
starting in 2007 up to a peak of 0.69 in 2010, and so does the
agricultural sector with the sharpest increase in 2008 when the average
pairwise correlation for the sector jumps to 0.30. The energy sector
also shows a substantial progression starting in 2007 and peaking at
0.63 in 2008 before levelling back down over the rest of the period. The
onset of the financial crisis seems therefore to have had an across the
board impact on the interdependence of commodities in contrast with
financialisation, much more closely related to the metals sector.

The entire cross-section sees an increase in average regression
coefficient relative to the market factor with the strongest progression
for the GB traded metals followed by the agricultural sector
(Table~\ref{tbl-regressions-index}). The pricing performance of the
factor dominates the US commodity complex
(Table~\ref{tbl-regressions-factors}), showing a 50\% increase in mean
average \(R^{2}\) to over 31\%, while that for both the CHP and open
interest factor decreases and, although increasing, remains modest for
the term structure factors; the increase is strongest for the GB traded
metal assets with an average \(R^{2}\) of almost 34\%. Taken together,
these results indicate the presence of a systematic factor across the
entire cross-section of the US commodities.

@@

The returns pull-back continues into the post-crisis period
(Table~\ref{tbl-stats-combined}) with twenty-two US and four GB traded
assets showing lower mean returns relative to the crisis while the same
number of US commodities and all GB metals show lower figures relative
to the financialisation period; mean returns for equally weighted
portfolios of US and GB assets drop down to -0.2\% and 2.7\%
respectively. Interestingly volatility also recedes substantially with
twenty US assets and all GB metals showing lower volatility figures
relative to both the crisis and financialisation periods suggesting that
investment outflows from the commodity futures complex have had a lower
impact on volatility than the corresponding inflows witnessed over the
financialisation period. The figures for the equally weighted portfolios
of US and GB assets drop to 9.9\% and 16.2\% respectively, very close to
pre-financialisation levels.\\
The pattern of reversion toward the pre-financialisation Keynesian
paradigm initiated during the crisis continues stronger into the
post-crisis period with mean returns higher during phases of
\textbf{backwardation} relative to \textbf{contango}, more so than over
the crisis period with seventeen US commodities showing higher mean
returns over \textbf{backwardation} phases. The difference is
significant at the 5\% level for Lumber (XCME) and at the 10\% level for
Palladium (XNYM) while no difference is significant for the assets
showing higher figures over phases of \textbf{contango}. The pattern is
more pronounced for the GB traded subset with all six assets showing
higher mean returns in \textbf{backwardation} with the difference
significant at the 1\% level for Nickel-primary (XLME), 5\% level for
Aluminium (XLME) and 10\% level for Zinc (XLME). It also extends to
equally weighted portfolios of US and GB traded assets that both show
higher mean returns during \textbf{backwardation} with the difference
significant at the 5\% level for the GB portfolio and the figure close
to the 10\% level significance threshold for the portfolio of US assets.
Volatilities are overall higher in \textbf{contango} although less so
than over the previous period with seventeen US commodities showing
higher figures over \textbf{contango} against twenty-three in the crisis
period. The pattern extends to three of the six GB traded metals with
the difference significant at the 1\% level for Copper (XLME) and Tin
(XLME) and at the 5\% level for Zinc (XLME) while differences are not
significant for Aluminium (XLME), Lead (XLME) and Nickel-primary (XLME)
that exhibit higher volatility over \textbf{backwardation} phases.
Equally weighted portfolios of US and GB traded assets on the other hand
both show higher volatility in \textbf{contango} with the difference
significant at the 1\% level for both. Over the financialisation period
in contrast volatilities were predominantly higher in
\textbf{backwardation} for US traded assets with fourteen commodities
showing higher volatility figures in these phases; the post-crisis
pattern of results for volatilities thus likewise shows signs of
reversion to pre-financialisation standards. After showing signs of
convergence in the crisis period the dynamics of the explanatory power
of CHP and \textbf{CHP} part ways (Table~\ref{tbl-time-reg-US-CHP}),
seemingly reverting to the Keynesian paradigm observed
pre-financialisation, further suggesting a return to legacy commodity
specific dynamics. The recovery initiated in the crisis period for that
of CHP is confirmed with a stronger increase, 27.2\% vs.~22.5\%,
strongest for the metals sector, 37.3\% vs.~23.7\%, while the energy
sector shows the opposite pattern: 11\% vs.~16\%. In contrast, after
increasing stronger in the crisis, that for \textbf{CHP} falls overall
post-crisis, 7.5\% vs 14.7\%, with the decrease strongest for the energy
sector in a relative sense: 2.5\% vs.~14.8\%.

The average pairwise correlations fall across the board, below the
levels observed during financialisation, with the figure for the entire
cross-section (Table~\ref{tbl-correlations-periods}) dropping down to
0.12 from 0.28 over the crisis. This substantial decline, that does not
seem to have been widely noted, appears to begin in the latter part of
the crisis period, around 2012 (Table~\ref{tbl-correlations-years}),
before levelling down over the whole post-crisis period.

Similarly, the average market factor beta for the entire cross-section
drops down to 0.94 (Table~\ref{tbl-regressions-index}), below the level
observed during financialization. The decline seems driven by the metals
sector, particularly the GB traded segment, with the figures for the US
and GB traded assets at 0.91 and 0.72, down from 1.12 and 1.18,
respectively. Meanwhile, the pricing performance of the market factor
also retreats to below financialisation levels with the average
\(R^{2}\) for the entire cross-section dropping down to 14.7\%
(Table~\ref{tbl-regressions-factors}). All sectors seem affected with
agriculturals, energy and metals all seeing their average \(R^{2}\)
decline below the levels observed during the financialisation period
albeit remaining above 10\% for the GB traded metals against less than
2\% pre-financialisation. Similarly, the performance for both the CHP
and term structure factors falls back to levels below those attained
over financialisation. These results, along with those from the crisis
period, suggest that the exogenous shock of the financial crisis,
combined with the subsequent accommodative monetary policy, may have
triggered a reversion toward pre-financialization fundamentals in the
commodity futures complex; an issue that, to our knowledge, has not been
widely discussed in the existing literature.\\
The phases analysis further documents the emerging role/dominance of the
\textbf{contango} regime post-crisis with average correlations higher
during those phases and the pattern stronger for the metal assets both
US and GB traded. Similarly, for the latter the pricing performance of
the market factor is considerably better during phase of
\textbf{contango} in contrast with the previous periods that showed no
substantial difference between phase of \textbf{contango} and
\textbf{backwardation}. The figures for the US and GB traded segments of
the sector are 23\% (\textbf{contango}) vs 11.5\%
(\textbf{backwardation}) against 36.2\% vs 43\% during financialisation
and 14.4\% (\textbf{contango}) vs 6.6\% (\textbf{backwardation}) against
33.5\% vs 34.4\% during financialisation respectively. Taken together,
these results indicate some recent changes in the correlation structure
within the sector and provide additional evidence that the
interdependence effects of financialization seems to have been strongest
for the entire cross-section of the metals sector with a lesser impact
for the other sectors. This observation, along with the decline in
interdependence in the commodity sector at large documented above, could
form a valuable basis for future research.

@@

\begin{longtable}[t]{>{}l>{}l>{}l>{}l>{}r>{}r}

\caption{\label{tbl-regime-diference-tests}This table shows the results
for tests of difference in mean (t-test) and variance (F-test) between
phases of aggregate backwardation and contango for the twenty four
individual US commodities and the six GB metals considered in the study
as well as for two equally weighted portfolios formed from the US
commodities and the GB metals respectively across the four periods of
interest (past: 1997-2003; financialisation: 2004-2008; crisis:
2008-2013; post-crisis: 2013-2018). 1\%, 5\% and 10\% p-value levels are
marked with ***, ** and * respectively in the significance column.
Aggregate CHP construction and corresponding regime definitions are
discussed in Section~\ref{sec-methods} while the results are discussed
in Section~\ref{sec-results}.}
\textcolor{black}{\textbf{asset}} & \textcolor{black}{\textbf{period}} & \textcolor{black}{\textbf{moment}} & \textcolor{black}{\textbf{dominant regime}} & \textcolor{black}{\textbf{p-value}} & \textcolor{black}{\textbf{significance}}\\
\textbf{\cellcolor{gray!10}{Corn-\#2 yellow (XCBT)}} & \textbf{\cellcolor{gray!10}{past}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{34.26\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Corn-\#2 yellow (XCBT)} & \textbf{past} & \textbf{volatility} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{89.69\%} & \textcolor[HTML]{4285f4}{}\\
\textbf{\cellcolor{gray!10}{Corn-\#2 yellow (XCBT)}} & \textbf{\cellcolor{gray!10}{financialisation}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{91.01\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Corn-\#2 yellow (XCBT)} & \textbf{financialisation} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{47.46\%} & \textcolor[HTML]{4285f4}{}\\
\textbf{\cellcolor{gray!10}{Corn-\#2 yellow (XCBT)}} & \textbf{\cellcolor{gray!10}{crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{40.83\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Corn-\#2 yellow (XCBT)} & \textbf{crisis} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{0\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Corn-\#2 yellow (XCBT)}} & \textbf{\cellcolor{gray!10}{post-crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{57.39\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Corn-\#2 yellow (XCBT)} & \textbf{post-crisis} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{5.02\%} & \textcolor[HTML]{4285f4}{*}\\
\textbf{\cellcolor{gray!10}{Oats (XCBT)}} & \textbf{\cellcolor{gray!10}{past}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{38.12\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Oats (XCBT)} & \textbf{past} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{0\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Oats (XCBT)}} & \textbf{\cellcolor{gray!10}{financialisation}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{35.29\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Oats (XCBT)} & \textbf{financialisation} & \textbf{volatility} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{0\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Oats (XCBT)}} & \textbf{\cellcolor{gray!10}{crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{93.66\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Oats (XCBT)} & \textbf{crisis} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{0\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Oats (XCBT)}} & \textbf{\cellcolor{gray!10}{post-crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{82.72\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Oats (XCBT)} & \textbf{post-crisis} & \textbf{volatility} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{0.03\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Soybean meal (XCBT)}} & \textbf{\cellcolor{gray!10}{past}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{66.39\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Soybean meal (XCBT)} & \textbf{past} & \textbf{volatility} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{0.12\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Soybean meal (XCBT)}} & \textbf{\cellcolor{gray!10}{financialisation}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{81.2\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Soybean meal (XCBT)} & \textbf{financialisation} & \textbf{volatility} & \textbf{backwardation} & \textcolor[HTML]{4285f4}{51.83\%} & \textcolor[HTML]{4285f4}{}\\
\textbf{\cellcolor{gray!10}{Soybean meal (XCBT)}} & \textbf{\cellcolor{gray!10}{crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{56.57\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Soybean meal (XCBT)} & \textbf{crisis} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{0\%} & \textcolor[HTML]{4285f4}{***}\\
\textbf{\cellcolor{gray!10}{Soybean meal (XCBT)}} & \textbf{\cellcolor{gray!10}{post-crisis}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{75.13\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Soybean meal (XCBT)} & \textbf{post-crisis} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{87.31\%} & \textcolor[HTML]{4285f4}{}\\
\textbf{\cellcolor{gray!10}{Soybean oil (XCBT)}} & \textbf{\cellcolor{gray!10}{past}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{backwardation}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{31.59\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
\textbf{Soybean oil (XCBT)} & \textbf{past} & \textbf{volatility} & \textbf{contango} & \textcolor[HTML]{4285f4}{81.72\%} & \textcolor[HTML]{4285f4}{}\\
\textbf{\cellcolor{gray!10}{Soybean oil (XCBT)}} & \textbf{\cellcolor{gray!10}{financialisation}} & \textbf{\cellcolor{gray!10}{mean}} & \textbf{\cellcolor{gray!10}{contango}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{92.18\%}} & \textcolor[HTML]{4285f4}{\cellcolor{gray!10}{}}\\
